Shapes analysis for time series.

### 2024-05-22 NEURIPS24/paper.tex
update

 \vspace{-1ex}
    \paragraph{Choice for the kernel associated with the RKHS $\msv$}
    \label{paragraph:kernel_V}
    As depicted on \Cref{fig:diffeo}-\ref{fig:transport}, we can not use any kernel $K$ to apply the previous methodology to learn deformations on time series' graphs.
    We describe and motivate our choice in this paragraph.
     Denote the one-dimensional Gaussian kernel by $K_\sigma^{(a)}(x,y)=\exp(-|x-y|^2/\sigma)$ for any $(x,y)\in (\Rset^a)^2$, $a\in \Nset$ and $\sigma>0$.
    To solve the geodesic shooting problem \eqref{eq:relaxation} on $\Rset^{d+1}$, we consider for $\msv$ the RKHS associated with the kernel defined for any $(t,x),(t',x')\in (\Rset^{d+1})^2$:
    \begin{align}
      \label{eq:kernel_TAS}
      &K_{\msg}((t,x),(t',x'))=\begin{pmatrix}
        c_0K_{\text{time}} & 0 \\
        0 & c_1 K_{\text{space}} 
        \end{pmatrix} \eqsp ,\\
       & K_{\text{space}}=K_{\sigma_{T,1}}^{(1)}(t,t')K_{\sigma_x}^{(d)}(x,x') \Idd\eqsp,K_{\text{time}}=K_{\sigma_{T,0}}^{(1)}(t,t') \eqsp,
    \end{align}
    parametrized by the widths $\sigma_{T,0},\sigma_{T,1},\sigma_x>0$ and the constants $c_0,c_1>0$.
This choice for $K_\msg$ is motivated by the representation \Cref{theorem:representation} and the following result. 
    \begin{lemma}
      \label{lemma:choice_of_kernel_V}
      If we denote by $\msv$ the RKHS associated with the kernel $K_{\msg}$, then for any vector field $v$ generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0},
       there exist $\gamma \in \msd(\Rset) $ and $f\in \rmC^1(\Rset^{d+1},\Rset^d)$ such that $\phi^v=\Psi_\gamma\circ\Phi_f $.
    \end{lemma}
    \sam{Parler des Cauchy kernel en appndice et du choix de la loss}
    % \begin{proof}
    %   Let $v$ be a vector field generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0}.
    %  We remark that the first coordinate of the velocity field $v_\tau$ denoted by $v_\tau^{\text{time}}$ only depends on the time variable $t$ for any $\tau\in[0,1]$.
    %  Thus, when computing the first coordinate of the deformation $\phi^v$, denoted by $\gamma$, we integrate \eqref{eq:LDDMM_dynamic} with $v_\tau$ replaced by $v_\tau^{\text{time}}$,
    %   thus $\gamma$ is independant of the variable $x$. Moreover, $\gamma\in \mcd(\Rset)$ since a Gaussian kernel induced an Hilbert space $\msv$ satisfying $|f|_V\leq |f|_\infty+ |\dd f|_\infty  $ for any $f\in \msv$ by \citep[Theorem 9]{glaunes2005transport}.
    %   For the same reason, we have $\phi^v\in \mcd(\Rset^{d+1})$, and thus its last coordinates denoted by $f$ belongs to $\rmC^1(\Rset^{d+1},\Rset^d)$, and by construction $\phi^v=\Psi_\gamma\circ\Phi_f $.
    % \end{proof
    \begin{remark}
      With this choice of kernel, the features associated to the time transformation can be extracted from the momentums $(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0}$ in \eqref{eq:def_v0} by taking the coordinates related to time.
      However, the features related to the space transformation are not only in the space coordinates since the related kernel $K_{\text{space}}$ depends on time as well.
     \end{remark}
    In \Cref{appendix:kernel_TS_LDDMM}, we give guidelines for selecting the hyperparameters $(\sigma_{T,0},\sigma_{T,1},\sigma_x,c_0,c_1)$.

@@

In practice, we set the kernel $k$ as the product of two anisotropic Gaussian kernels, $k_{\pos}$ and $k_{\dir}$, 
such that for any $(x,\overrightarrow{u}),(y,\overrightarrow{v}) \in (\Rset^{d+1} \times \mathbb{S}^d)^2$
\begin{equation}
    k((x,\overrightarrow{u}),(y,\overrightarrow{v})) = k_{\pos}(x,y)k_{\dir}(\overrightarrow{u},\overrightarrow{v}) \eqsp.
  \end{equation}
  The specific kernels $k_{\pos},k_{\dir}$ that we use in our experiments are given \Cref{appendix:kernel_implementation}.
   Note that the loss kernel $k$ has nothing to do with the velocity field kernel denoted by $K_\msg$ or $K$ specified in \Cref{paragraph:kernel_V}.
Finally, we define the data fidelity loss function, $\scrl$, as $ d_{\msw^*}^2$, which is differentiable with regards to its first variable.
For further readings on curves and surfaces representation as varifolds, readers can refer to \cite{kaltenmark2017general,charon2013varifold}. 
{\color{red} mettre une phrase sur la loss multiscale}

@@

  All these state-of-the-art methods previously mentionned \cite{glaunes2008large,oh2024stable,wu2024shape,heo2024logistic} are comapred to TS-LDDMM in \Cref{appendix:classification_shape_analysis}.
  \section{Limitations}
  {\color{red} rappeler les limites de la méthodes : petite dimension à cause des kernels, tuner les hyper paramètres correctement, version adaptative}
\section{Conclusion}
\vspace{-1ex}
In this paper, we propose a feature representation method, TS-LDDMM, designed for 
shape comparison in homogeneous time series datasets. We show on a real dataset 
its ability to study, with high interpretability, the inter-individual shape 
variability. As an unsupervised approach, it is user-friendly and enables knowledge 
transfer for different supervised tasks such as classification. Although TS-LDDMM 
is already competitive for classification, its performances can be leveraged on 
more heterogeneous datasets using a hierarchical clustering extension, which is relagated for future work.

>>>>

 \vspace{-1ex}
    \paragraph{Choice for the kernel associated with the RKHS $\msv$}
    \label{paragraph:kernel_V}
    As depicted on \Cref{fig:diffeo}-\ref{fig:transport}, we can not use any kernel $K$ to apply the previous methodology to learn deformations on time series' graphs.
    We describe and motivate our choice in this paragraph.
     Denote the one-dimensional Gaussian kernel by $K_\sigma^{(a)}(x,y)=\exp(-|x-y|^2/\sigma)$ for any $(x,y)\in (\Rset^a)^2$, $a\in \Nset$ and $\sigma>0$.
    To solve the geodesic shooting problem \eqref{eq:relaxation} on $\Rset^{d+1}$, we consider for $\msv$ the RKHS associated with the kernel defined for any $(t,x),(t',x')\in (\Rset^{d+1})^2$:
    \begin{align}
      \label{eq:kernel_TAS}
      &K_{\msg}((t,x),(t',x'))=\begin{pmatrix}
        c_0K_{\text{time}} & 0 \\
        0 & c_1 K_{\text{space}} 
        \end{pmatrix} \eqsp ,\\
       & K_{\text{space}}=K_{\sigma_{T,1}}^{(1)}(t,t')K_{\sigma_x}^{(d)}(x,x') \Idd\eqsp,K_{\text{time}}=K_{\sigma_{T,0}}^{(1)}(t,t') \eqsp,
    \end{align}
    parametrized by the widths $\sigma_{T,0},\sigma_{T,1},\sigma_x>0$ and the constants $c_0,c_1>0$.
This choice for $K_\msg$ is motivated by the representation \Cref{theorem:representation} and the following result. 
    \begin{lemma}
      \label{lemma:choice_of_kernel_V}
      If we denote by $\msv$ the RKHS associated with the kernel $K_{\msg}$, then for any vector field $v$ generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0},
       there exist $\gamma \in \msd(\Rset) $ and $f\in \rmC^1(\Rset^{d+1},\Rset^d)$ such that $\phi^v=\Psi_\gamma\circ\Phi_f $.
    \end{lemma}
    Instead of Gaussian kernels, other types of smooth kernel can be selected as long as the structure \eqref{eq:kernel_TAS} is respected.
    % \begin{proof}
    %   Let $v$ be a vector field generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0}.
    %  We remark that the first coordinate of the velocity field $v_\tau$ denoted by $v_\tau^{\text{time}}$ only depends on the time variable $t$ for any $\tau\in[0,1]$.
    %  Thus, when computing the first coordinate of the deformation $\phi^v$, denoted by $\gamma$, we integrate \eqref{eq:LDDMM_dynamic} with $v_\tau$ replaced by $v_\tau^{\text{time}}$,
    %   thus $\gamma$ is independant of the variable $x$. Moreover, $\gamma\in \mcd(\Rset)$ since a Gaussian kernel induced an Hilbert space $\msv$ satisfying $|f|_V\leq |f|_\infty+ |\dd f|_\infty  $ for any $f\in \msv$ by \citep[Theorem 9]{glaunes2005transport}.
    %   For the same reason, we have $\phi^v\in \mcd(\Rset^{d+1})$, and thus its last coordinates denoted by $f$ belongs to $\rmC^1(\Rset^{d+1},\Rset^d)$, and by construction $\phi^v=\Psi_\gamma\circ\Phi_f $.
    % \end{proof
    \begin{remark}
      With this choice of kernel, the features associated to the time transformation can be extracted from the momentums $(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0}$ in \eqref{eq:def_v0} by taking the coordinates related to time.
      However, the features related to the space transformation are not only in the space coordinates since the related kernel $K_{\text{space}}$ depends on time as well.
     \end{remark}
    In \Cref{appendix:kernel_TS_LDDMM}, we give guidelines for selecting the hyperparameters $(\sigma_{T,0},\sigma_{T,1},\sigma_x,c_0,c_1)$.

@@

In practice, we set the kernel $k$ as the product of two anisotropic Gaussian kernels, $k_{\pos}$ and $k_{\dir}$, 
such that for any $(x,\overrightarrow{u}),(y,\overrightarrow{v}) \in (\Rset^{d+1} \times \mathbb{S}^d)^2$
\begin{equation}
    k((x,\overrightarrow{u}),(y,\overrightarrow{v})) = k_{\pos}(x,y)k_{\dir}(\overrightarrow{u},\overrightarrow{v}) \eqsp.
  \end{equation}
   Note that the loss kernel $k$ has nothing to do with the velocity field kernel denoted by $K_\msg$ or $K$ specified in \Cref{paragraph:kernel_V}.
Finally, we define the data fidelity loss function, $\scrl$, as a sum of $ d_{\msw^*}^2$ using different kernel's width parameters $\sigma$ to incorporate multiscale information such that. $\scrl$ is indeed differentiable with regards to its first variable.
The specific kernels $k_{\pos},k_{\dir}$ that we use in our experiments are given \Cref{appendix:kernel_implementation}.
For further readings on curves and surfaces representation as varifolds, readers can refer to \cite{kaltenmark2017general,charon2013varifold}.

@@

  All these state-of-the-art methods previously mentionned \cite{glaunes2008large,oh2024stable,wu2024shape,heo2024logistic} are comapred to TS-LDDMM in \Cref{appendix:classification_shape_analysis}.
  \section{Limitations}
  While TS-LDDMM performs well on shape-based datasets, it assumes a certain degree of homogeneity with the existence of a reference time series graph$\msg_0$.
  Extending TS-LDDMM to more heterogeneous datasets using clustering is ongoing work.
   TS-LDDMM employs kernel computations, which require specific libraries (e.g., KeOps \cite{charlier2021kernel}) to be efficient and scalable.
   However, in our experiments, the time complexity of TS-LDDMM is comparable to that of competitors.
   It is clear that TS-LDDMM needs to be extended to handle very large datasets with high-dimensional time series (such as videos).
    Additionally, TS-LDDMM requires tuning several hyperparameters, though this is a common requirement among competitors \cite{glaunes2008large, oh2024stable, wu2024shape, heo2024logistic}.
    In future work, adaptive methods are expected to be developed to provide a more user-friendly interface.
  {%\color{red} rappeler les limites de la méthodes : petite dimension à cause des kernels, tuner les hyper paramètres correctement, version adaptative}
\section{Conclusion}
\vspace{-1ex}
In this paper, we propose a feature representation method, TS-LDDMM, designed for 
shape comparison in homogeneous time series datasets. We show on a real dataset 
its ability to study, with high interpretability, the inter-individual shape 
variability. As an unsupervised approach, it is user-friendly and enables knowledge 
transfer for different supervised tasks such as classification. Although TS-LDDMM 
is already competitive for classification, its performances can be leveraged on 
more heterogeneous datasets using a hierarchical clustering extension, which is relagated for future work.
